---
title: "Choice alternatives"
description: >
  How to define choice alternatives?
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choice alternatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# library("choicedata")
devtools::load_all() # TODO: remove
```

The choice alternatives form the choice set from which the decision makers select [@Train2009]. In the `{choicedata}` package, these alternatives are represented as a `choice_alternatives` S3 object [@Chambers1992].  

For instance, consider households choosing between different heating fuels, a topic extensively studied for energy use forecasting and climate change analysis. The choice alternatives in this scenario could include gas, electricity, and oil. The corresponding `choice_alternatives` object can be constructed as follows:

```{r, choice alternatives example 1}
choice_alternatives(
  J = 3,
  alternatives = c("gas", "electricity", "oil"),
  base = "gas"
)
```

Here, 

- `J` denotes the total number of alternatives.

- The `alternatives` argument can optionally be used to label the alternatives. Internally, they will be sorted alphabetically.

<!-- TODO: link for "choice formula" -->

- The `base` argument indicates the base alternative for covariates that are not alternative-specific, as detailed in the vignette on the choice formula. When printing the `choice_alternatives` object, the base alternative is marked with a `*`.

<!-- TODO: links for "choice data" and "choice probabilities" -->

Once a `choice_alternatives` object has been specified, it can be used to simulate choice data or compute choice probabilities.

<!-- TODO: general example -->

## References
